---
title: "The Exponential Distribution in R and the Central Limit Theorem"
author: "James Sheldon"
date: "May 23, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r system}
sessionInfo()
```
## Overview

This is part 1 of the final course project for the statistical inference class from Coursera presented by Johns Hopkins University. The purpose of this exercise is to explore how the Exponential Distribution relates to the Central Limit Theorem using one thousand mean simulations.

## Simulation

```{r}
set.seed(10736)
lambda <- 0.2 # fixed rate parameter
n <- 40 # number of exponentials
sim <- 1000 # fixed number of simulations per instruction

# the exponential distribution
plot(rexp(10000,lambda), pch=20, cex=0.6, main="The Exponential Distribution for 10,000 Observations")
```

```{r}
# generate the collection of means for 1,000 simulations of the exponential distribution
means <- NULL
for (i in 1 : sim) means <- c(means, mean(rexp(n,lambda)))
hist(means, col="blue", main="rexp Mean Mistribution", breaks=40)
rug(means)
```

## Sample Mean Versus Theoretical Mean

Since the theoretical mean of the exponential distribution is 1/lambda and in our examples we are using lambda = 0.2, the sample and theoretical mean should both be at or near 5.

```{r}
hist(means, col="blue", main="Actual vs Theoretical Mean", breaks=20)
abline(v=mean(means), lwd="3", col="red")
text(6.1, 90, paste("Actual Mean = ", round(mean(means),2), "\n Theoretical Mean = 5" ), col="red")
```

It looks like we're in pretty good shape.

## Sample Variance Versus Theoretical Variance

Since the standard deviation of the exponential distribution is (1/lambda)/sqrt(n) we see the following for the standard deviation:

```{r}
print(paste("Theoretical: ", round((1/lambda)/sqrt(n),4), ", Sample", round(sd(means),4)))
```

and for the variances we have

```{r}
print(paste("Theoretical: ", (1/lambda)^2/n, ", Sample", round(var(means),4)))
```

Based on the coursework, this suggests we're on track with the theoretical values.

## Is the Distribution of Means Normal?

Finally we were asked to explore whether or not the exponential distribution is close to normal. It certainly would seem as though it should be based on the Central Limit Theorem.

```{r}
hist(means, prob=TRUE, col="blue", main="Mean Distribution for rexp()", breaks=20)
lines(density(means), lwd=3, col="red")
```

Based on this graph it certainly seems as though the theory matches up with our work.