---
title: "Tooth Growth Analysis"
author: "James Sheldon"
date: "May 24, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(datasets)
library(graphics)
```
```{r system}
sessionInfo()
```

## Overview

This analysis pertains to the ToothGrowth dataset made available in the R platform.  We will load and summarize the data, perform some exploratory analysis and use confidence intervals and hypothesis tests to determine any relationships between tooth growth and the other variables.

```{r teeth}
data("ToothGrowth")
summary(ToothGrowth)
str(ToothGrowth)
head(ToothGrowth)
tail(ToothGrowth)
unique(ToothGrowth$dose)
unique(ToothGrowth$supp)
by(ToothGrowth$len, INDICES = list(ToothGrowth$supp, ToothGrowth$dose), summary)
```

What we're seeing is this data represents the delivery of three different levels of Vitamin C (0.5, 1, and 2 mg) via two different vectors:  orange juice or ascorbic acid.  There are 60 total observations with three variables:

1. `len` which is a numeric value that represents the tooth length.
2. `supp` is the delivery vector.
3. `dose` is the numeric measure of the dose size in milligrams.

## Exploratory Analysis

Let's now explore the data a bit.  We'll take a look at a number of boxplots.  This should hopefully help us at least get an idea on whether or not there are correlations between different data elements.  Of course, it's important to remember that *correlation does not imply causation.* We are merely looking for some correlations and will explore those correlations further once they're noted.



```{r tooth supp and dose, echo=TRUE}
avg <- aggregate(len~.,data=ToothGrowth,mean)
#Plot the Tooth Length relative to Dosage & Supplement
ggplot(data = ToothGrowth,aes(x=dose,y=len)) + geom_point(aes(group=supp,colour=supp,size=2,alpha=0.6)) + geom_line(data=avg,aes(group=supp,colour=supp)) + labs(title="Tooth Length vs Dosage & Supplement")
```

```{r tooth and supp}
ggplot(aes(x = supp, y = len), data = ToothGrowth) + geom_boxplot(aes(fill = supp)) + labs(title="Tooth Length sv Supplement")
```

```{r tooth and dose}
ggplot(aes(x = factor(dose), y = len), data = ToothGrowth) + geom_boxplot(aes(fill = factor(dose))) + labs(title="Tooth length vs Dosage")
```

```{r}
ggplot(aes(x = supp, y = len), data = ToothGrowth) + geom_boxplot(aes(fill = supp)) + facet_wrap(~ dose) + labs(title="Tooth Length vs Supplement & Dosage")
```

The above plots illustrate there is a strong positive correlation between dosage and tooth length.  on the surface, the vector being employed doesn't show as strong of a correlation; however, it is too early to rule that out.

## Confidence Testing

Now let's apply some t-tests to determine confidence intervals for any correlations we may be seeing in the data.

```{r welch1}
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = ToothGrowth)
```


```{r welch2}
dose1 <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
dose2 <- subset(ToothGrowth, dose %in% c(0.5, 2.0))
dose3 <- subset(ToothGrowth, dose %in% c(1.0, 2.0))
t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = dose1)
```


```{r}
t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = dose2)
```

```{r}
t.test(len ~ dose, paired = FALSE, var.equal = FALSE, data = dose3)
```

```{r}
Tooth.dose4 <- subset(ToothGrowth, dose == 0.5)
Tooth.dose5 <- subset(ToothGrowth, dose == 1.0)
Tooth.dose6 <- subset(ToothGrowth, dose == 2.0)
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = Tooth.dose4)
```

```{r}
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = Tooth.dose5)
```

```{r}
t.test(len ~ supp, paired = FALSE, var.equal = FALSE, data = Tooth.dose6)
```

## Conclusions

From the above confidence interval results we can conclude that there is a significant correlation between tooth length and dose levels.
